---
title: "Metabolomic_Analysis"
author: "Jacob Agerbo Rasmussen"
date: "7/16/2020"
output:
  html_document: default
  pdf_document: default
---
All the data to reproduce analysis can be found here:
(https://github.com/JacobAgerbo/Rasmussen_et_al_2020a/)

Initial inspection of the reads to identify and remove contaminants.
Plotting of the negative controls composition.
```{r load dependencies}
library(ggplot2)
library(readxl)
library(phyloseq)
library(ape)
library(dplyr)
library(MetaboDiff)
library(RColorBrewer)
library(cowplot)
library(EnhancedVolcano)
library(ggpubr)
library(rstatix)
library(beeswarm)
library(cluster)
```

```{r load data}
# Load data
ft <- read_excel("featuretable.xlsx")
tax <- read_excel("Chemical_Class.xlsx")
md <- read_excel("metadata.xlsx")
MetDNA <- read.csv("HF_pseudomonas_MRN.annotation.result.csv")

My_pal <- c("#3889A0","#F2AD00","#9C964A","#D95F02","#1B9E77","#046C9A","#0B775E","#35274A","#F2300F","#666666") 
group_pal <- c("#e3aa74", "#ed828c", "#7bb6bd")
```

To minimise false positives we included procedural blanks in the data generation. Metabolites which are present in the blanks should be removed (at least to my believe).
To include a little bit of plasticity in filtering of blanks i include only metabolites, which are lower than 5*10^6 in blanks.
```{r remove metabolites present in blanks}
## Remove features present in blanks
blk_ids <- which(colnames(ft) %in% as.character(md$SampleID[grep("BLANK",md$SampleType )]))
blanks <- ft[,blk_ids]
blanks <- blanks[apply(blanks[,-1], 1, function(x) !all(x < 1*10^5)),]
blank_ids <- rownames(blanks)
ft.filtered <- ft[,-c(blk_ids)]
ft.filtered = ft.filtered[-which(rownames(ft.filtered) %in% blank_ids), ]
# remove QCs
QC_ids <- which(colnames(ft.filtered) %in% as.character(md$SampleID[grep("QC",md$SampleType )]))
ft.filtered_all <- ft.filtered[,-c(QC_ids)]
ft <- ft.filtered_all
# set ID as rownames 
ft <- as.matrix(ft)
rownames(ft) <- ft[,1]
ft <- ft[,-c(1)]
```

To minimise highly zero inflated dataset, I remove metabolites which are present in less than 50 of the 60 samples.
```{r remove metabolites present in blanks}
# Remove metabolites not present in 50 of 60 of the samples
ft_curated <- ft[apply(ft, 1, FUN = function(x){sum(x == 0)}) < 10,]
# post blank removal
dim(ft)
# post inflation removal
dim(ft_curated)
ft <- ft_curated
```

```{r Curate tax adn md files}
## check if datasets are identically ordered
## Due to a bug, there is one metabolite no. "124151", which is NA in tax
ft <- ft[rownames(ft) != "125141",]
## Filter tax and MetDNA annotation files of remove metabolites
tax <- tax[match(rownames(ft),tax$ID),]
MetDNA <- MetDNA[match(rownames(ft),MetDNA$name),]

#combine tax and MetDna
Annotation <- as.data.frame(tax)

cmp <- MetDNA$compound.name
Annotation$Annotation <- paste(Annotation$Compound_Name,cmp)
## remove "NA NA" and make it to NA, so pasted text looks better
Annotation$Annotation <- gsub("NA NA", NA, as.character(Annotation$Annotation))
Annotation$Annotation <- gsub("NA ", "", as.character(Annotation$Annotation))
Annotation$Annotation <- gsub(" NA", "", as.character(Annotation$Annotation))

identical(tax$ID,as.numeric(rownames(ft)))
identical(MetDNA$name,as.integer(rownames(ft)))
identical(Annotation$ID,as.numeric(rownames(ft)))
## Cool seems to be identical

#Filter out blanks and QC from metadata file
md <- md[match(colnames(ft),md$SampleID),]
rownames(md) <- md$SampleID
identical(rownames(md),colnames(ft))
rownames(md) <- colnames(ft) # to minimise difficulties with phyloseq
md <- as.matrix(md) # to minimise difficulties with phyloseq
md <- as.data.frame(md) # to minimise difficulties with phyloseq - i dont know what is going on?! Probably a bug with class in rownames
```

I tend to clean up my working space, because i'm one of those who get lost in my own r codes....
```{r Clean up - some peepz just hate the mess...}
rm("blanks", "ft.filtered","ft.filtered_all", 
   "blank_ids", "blk_ids", "QC_ids","ft_curated", "cmp")
```

```{r generate phyloseq elements, because phyloseq is genious for this type of abundance data}
# physeq element
# since Phyloseq likes taxa to be in matrices, i will make data to matrix
Ant <- as.matrix(Annotation)
rownames(Ant) <- rownames(ft)

# combine procedural replicates 
#agg = aggregate(t(ft),
#                by = list(md$Individuals),
#                FUN = sum)
#rownames(agg) <- agg$Group.1
#agg <- agg[,-c(1)]
#ft <- t(agg)

## Make metadata compatible to metabolite data
#md <- md[md$RUN == "B",]
#colnames(ft) <- md$SampleID

#make phyloseq object for ordination
physeq <- phyloseq(otu_table(ft, taxa_are_rows = TRUE), tax_table(Ant),sample_data(md))
# sum normalise physeq object for ranked analysis
physeq.norm = transform_sample_counts(physeq, function(x) x/sum(x))
#make dataset only for metabolites versus mycoplasma
Myco_x_meta = subset_samples(physeq, Keep=="Yes")
```
\newpage
We start with a simple rank-abundance barplot, using the cumulative fractional abundance of each metabolite in the dataset.
This really visualise the sparsity of the dataset. Simply a lot of metabolites, which are very low abundant.
```{r barplot stuff,message=FALSE}
par(mar = c(10, 4, 4, 2) + 0.1) # make more room on bottom margin
barplot(sort(taxa_sums(physeq.norm), TRUE)[1:740]/nsamples(physeq.norm), las=2, col = "#1B9E77")
```
\newpage
```{r Class specific Ordination}
GP.ord <- ordinate(Myco_x_meta, "PCoA", "jaccard")
p2 = plot_ordination(Myco_x_meta, GP.ord, type="samples", color="Feed", axes = c(1,2)) 
md_ord <- md[md$Keep == "Yes",]
p_ALL =p2 + geom_point(size=7.5, alpha = as.numeric(md_ord$Mycoplasma)) + 
  theme_minimal() + scale_fill_manual(values = group_pal) +
  scale_color_manual(values = group_pal)



Myco_x_meta_Class <- subset_taxa(Myco_x_meta, CF_subclass=="Amino acids, peptides, and analogues")
GP.ord <- ordinate(Myco_x_meta_Class, "PCoA", "jaccard")
p2 = plot_ordination(Myco_x_meta_Class, GP.ord, type="samples", color="Feed", axes = c(1,2)) 
md_ord <- md[md$Keep == "Yes",]
p_AA =p2 + geom_point(size=7.5, alpha = as.numeric(md_ord$Mycoplasma)) + 
  theme_minimal() + scale_fill_manual(values = group_pal) +
  scale_color_manual(values = group_pal)

Myco_x_meta_Class <- subset_taxa(Myco_x_meta, CF_subclass=="Monoterpenoids" | CF_subclass=="Diterpenoids" |CF_subclass=="Triterpenoids" | CF_subclass=="Sesquiterpenoids")
GP.ord <- ordinate(Myco_x_meta_Class, "PCoA", "jaccard")
p2 = plot_ordination(Myco_x_meta_Class, GP.ord, type="samples", color="Feed", axes = c(1,3)) 
md_ord <- md[md$Keep == "Yes",]
p_terp =p2 + geom_point(size=7.5, alpha = as.numeric(md_ord$Mycoplasma)) + 
  theme_minimal() + scale_fill_manual(values = group_pal) +
  scale_color_manual(values = group_pal)

Myco_x_meta_Class <- subset_taxa(Myco_x_meta, CF_subclass=="Bile acids, alcohols and derivatives")
GP.ord <- ordinate(Myco_x_meta_Class, "PCoA", "jaccard")
p2 = plot_ordination(Myco_x_meta_Class, GP.ord, type="samples", color="Feed", axes = c(1,2)) 
md_ord <- md[md$Keep == "Yes",]
p_Bile =p2 + geom_point(size=7.5, alpha = as.numeric(md_ord$Mycoplasma)) + 
  theme_minimal() + scale_fill_manual(values = group_pal) +
  scale_color_manual(values = group_pal)

#pdf("jaccard_ordination_metabolites_x_mycoplasma_classes.pdf", height = 8, width = 12)
cowplot::plot_grid(p_ALL,
                   p_AA,
                   p_terp,                   
                   p_Bile,
                   labels = c('All Classes of metabolites','Amino acids, peptides, and analogues', 'Terpenoids',
                              'Bile acids, alcohols and derivatives'), ncol = 2)
#dev.off()
```
```{r Clean up - some peepz just hate the mess...}
rm("Myco_x_meta_Class","md_ord", 
   "MetDNA", "tax","physeq.norm", "p_Bile", "p_FA","p_AA","p_terp","p2")
```

\newpage
########################################################################
##    differential intensity of metabolites across samples            ##
########################################################################
We use metabodiff to analyse the dataset, since it is includes knn imputation and SVN normalisation (Mock et al. 2015). 
See also: https://github.com/andreasmock/MetaboDiff. 
```{r - }
met <- create_mae(ft,Ant,md)

met = knn_impute(met,cutoff=0.4)
met <- normalize_met(met)
quality_plot(met,
             group_factor="Feed",
             label_colors=group_pal)
```
```{r - Metabodiff diff test}
met = diff_test(met,
                group_factors = c("Feed"))
str(metadata(met), max.level=2)
```
```{r plot differential test, using Enhanced volcano}
# Extract data from test to use for selection of significantly different metabolites for heatmap
df <- met@metadata[["anova_Feed_Control_vs_Probiotics_vs_Synbiotics"]]
df$Metabolites <- as.integer(rownames(met@ExperimentList@listData[["raw"]]@assays@data@listData[[1]]))
tax_diff <-  Ant[order(match(df$Metabolites,rownames(Ant))),]
df <- cbind(df,tax_diff)

# Intercept is CTRL -----> PRO & SYN
p1 = EnhancedVolcano(df,
                     lab = c(ifelse(df$adj_pval<10e-3, 
                                    ifelse(df$CF_class != "no matches", 
                                           df$CF_class, ""), "")),
                     boxedLabels = TRUE,
                     col = c('grey0', 'grey0', 'red3', 'red3'),
                     x = 'fold_change',
                     y = 'adj_pval',
                     xlim = c(-10,10),
                     pCutoff = 0.05,
                     FCcutoff = 1.0,
                     labSize = 3,
                     pointSize = c(ifelse(df$adj_pval<10e-3, 5, 1)),
                     title = 'Differential Abundance of metabolites related to feed',
                     subtitle = 'Comparison between CTRL, PRO, and SYN',
                     xlab = bquote(~Log[2]~ 'Fold Change'),
                     .legend = c('NS','P & Log2 FC'),
                     legendLabels = c('NS',expression(p-value~and~log[2]~FC)),
                     legendPosition = 'bottom')

pdf("Differential Abundance of metabolites MetaboDiff template.pdf", height = 7.5, width = 15)
cowplot::plot_grid(p1, ncol = 1)
dev.off()

#write.csv(df,file = "Diff_test_Metabodiff.csv")
#write.csv(df_sig,file = "Sig_Diff_test_Metabodiff.csv")
```
\newpage
########################################################################
##    hierichal clustering of metabolites across samples              ##
########################################################################

```{r}
#column based normalisation prior comparison
hm.ft <- met@ExperimentList@listData[["norm"]]@assays@data@listData[[1]] # We use non imputated normalised data for heatmap
hm.md <- md #[md$Keep == "Yes",]
hm.md$Mycoplasma <- as.numeric(hm.md$Mycoplasma)

hm.ft <- as.data.frame(hm.ft)
# Select only significant different metabolites
df_sig <- df[df$pval<0.001,]
CTRL_sig <- df_sig[df_sig$fold_change < 0,]

names <- as.numeric(df_sig$ID)
hm.ft <- hm.ft[match(names,rownames(hm.ft)),]


# Make mean and SD for visualising heatmap
scaled.dat <- scale(t(hm.ft))
scaled.dat <- t(scaled.dat)
hm.ft <- scaled.dat
# configure data to log10, add 1 to remove zero intensity, which will conflict log10

# Get annotation for heatmap
hm.Annotation <- Annotation[match(rownames(hm.ft), Annotation$ID),]

### Plotting
#Clustering of PAMs
set.seed(2)
pa_r = pam(hm.ft, k = 6)
pa_c = pam(t(hm.ft), k = 2)

# Define some graphics to display the distribution of columns
.density = anno_density(hm.ft, type = "line", gp = gpar(col = "black", fill = "thistle" ))
col = list(Feed = c("Control" = "#e3aa74", "Probiotics" = "#ed828c", "Synbiotics" = "#7bb6bd"),
           PAM = c("1" = "#3889A0","2" = "#F2AD00","3" = "#f5452a","4" = "#D95F02", "5" = "#1B9E77" ,"6" = "#046C9A" ,"7" = "#0B775E" ,"8" = "#35274A" ,"9" = "#F2300F" ,"10" = "#666666"),
           Mycoplasma = circlize::colorRamp2(c(0, 1), c("white", "#1B9E77")))
# Annotate side graphs in heatmap
ha <- HeatmapAnnotation(
  Feed = hm.md$Feed, 
  Mycoplasma = hm.md$Mycoplasma,
  density = .density,height = unit(2, "cm"),
  col = col
)
ha_mix_right = HeatmapAnnotation(PAM = pa_r$clustering, which = "row", width = unit(1.5, "cm"),
                                 col = col)

#pdf("Metabolite_heatmap.pdf", width = 15, height = 10)
Heatmap(hm.ft, name = paste("VSN Metabolite","SD of Intensity", sep = "\n"), col <- c("#082ca3","#f6ffed","#ad1515"),
        top_annotation = ha,
        split = paste0("pam", pa_r$clustering),
        column_split = paste0("", pa_c$clustering), 
        row_labels = c(ifelse(hm.Annotation$CF_class != "no matches", hm.Annotation$CF_class, "")),
        column_labels = hm.md$Individuals,
        row_dend_side = "right",
        row_names_side = c("left"),
        width = 1) + ha_mix_right

#dev.off()
```
